
```{r include=FALSE}
eval_pop = FALSE
eval_IHME = FALSE
knitr::opts_chunk$set(echo = TRUE)
```

# Data

In this section we are going to download and explore datasets before editing and storing data into `duckdb`. Data sets consists of the population and disease prevalances information. It can be downloaded as (several) CSV-files from the links below:

* [Population data](http://ghdx.healthdata.org/record/ihme-data/gbd-2019-population-estimates-1950-2019)
* [IHME data](http://ghdx.healthdata.org/gbd-results-tool?params=gbd-api-2019-permalink/e77f5cf5b1b5b77568d115f4878cd19b)

Download data manually and store them to `data/` folder. There are many csv-files so loading those can be done easily with `vroom` and `fs` packages. Before transferring datasets into mighty `duckdb` database, we find out little bit more what these data sets holds. 

## Population

We are interested of the population by age and gender in European countries. 

### Load data into R

Load the `vroom()` library. Load one data set into `IHME2010` object and use the `id` argument to add the file name to the data frame.  Use **file_name** as the argument's value. Load the file spec into a variable called `spec`, using the `spec()` command.

```{r, eval = eval_pop, message=FALSE}
library(vroom)
IHME2010 <- vroom("data/IHME_GBD_2019_POP_2010_2019_SYA/IHME_GBD_2019_POP_SYA_2010_Y2021M01D28.CSV", id = "file_name")
spec <- spec(IHME2010)
spec
```

### Explore dataset

Explore dataset and it's variables by `summary()`

```{r, eval = eval_pop}
summary(IHME2010)
rm(IHME2010) # Remove dataset from memory
```


### Load and modify multiple files

Load the `fs` and `dplyr` libraries. Load to a variable named `files` list files in the **./data/** folder using the `dir_ls()` function, use the `glob` argument to pass a wildcard to list CSV files only. 

```{r, eval = eval_pop}
library(fs)
library(dplyr)
loc <- ("data/IHME_GBD_2019_POP_2010_2019_SYA/")
files <- dir_ls(loc, glob = "*.CSV")
files
```

Group `population` by `location_name, year_id, sex_name, age_group_name, file_name` and get the total of `val` and the number of items. Name them `total_population` and `items` respectively. Name the new dataset `countries`. Define the `countries` variable as `NULL` prior to the for loop and add a `bind_rows()` step to `countries` to preserve each summarized view. Remove the `population` variable at the end of each cycle.

```{r, eval = eval_pop, message=FALSE}
countries <- NULL
for(i in seq_along(files)) {
  population <- vroom(files[i], id = "file_name")
  
  countries <- population %>% 
    group_by(location_name, year_id, sex_name, age_group_name, age_group_id) %>% 
    summarise(total_population = sum(val, na.rm = T)) %>% 
    bind_rows(countries)
  
  rm(population)
}
```

Preview the `countries` dataset by checking how many countries are in the dataset and calculating total population by year

```{r, eval = eval_pop}
## How many countries/locations are included?    
length(unique(countries$location_name))

## Calculate countries population by year 
countries %>% 
  filter(sex_name == "both") %>% 
  group_by(location_name, year_id, sex_name) %>% 
  summarise(pop = sum(total_population)) -> countries_population

head(countries_population)
```

### Included European countries

Below is a list of 51 European countries. See which of these countries are NOT included in population dataset

```{r, eval = eval_pop}
europe <- c("Albania","Andorra","Armenia","Austria","Azerbaijan","Belarus","Belgium","Bosnia and Herzegovina","Bulgaria","Croatia","Cyprus","Czechia","Denmark","Estonia","Finland","France","Georgia","Germany","Greece","Hungary","Iceland","Ireland","Italy","Kazakhstan","Latvia","Lithuania","Luxembourg","Malta","Republic of Moldova","Monaco","Montenegro","Netherlands","North Macedonia","Norway","Poland","Portugal","Romania","Russian Federation","San Marino","Serbia","Slovakia","Slovenia","Spain","Sweden","Switzerland","Turkey","Ukraine","United Kingdom")
```

Create new dataset which has only europe countries and compare which are not included

```{r, eval = eval_pop}
eucountries <- countries %>% 
  filter(location_name %in% europe)

## How many European countries are included in the dataset?
length(unique(eucountries$location_name))

## These countries are missing
europe[!europe %in% eucountries$location_name]
```

### Europe total population in 2019

Calculate Total population in 2019 of selected European countries

```{r, eval = eval_pop}
eucountries %>% 
  filter(sex_name == "both" & year_id == 2019 & location_name %in% europe) %>% 
  group_by(2019) %>%
  summarise(pop = sum(total_population)) -> total_population_2019

total_population_2019
```


### PLOT: Finland population 2019

Draw a plot of Finland population in 2019 by age and gender

```{r, eval = eval_pop}
## re-adjust age_group variable of the population for filtering
eucountries$age_group <- eucountries$age_group_name
eucountries$age_group[eucountries$age_group == "<1 year"] <- 0
eucountries$age_group[eucountries$age_group == "95 plus"] <- 95
eucountries$age_group <- as.integer(eucountries$age_group)

## plot age distribution
library(ggplot2)
ggplot(data = eucountries[eucountries$sex_name %in% c("female", "male") & eucountries$location_name == "Finland" & eucountries$year_id == 2019, ]) +
  geom_bar(aes(x=age_group, y=total_population, fill=sex_name), stat = "identity") + 
  labs(x="age", y="freq", title="Finland population 2019")
```


### PLOT: European countries population

Filter and group dataset to get total population by years and countries

```{r, eval = eval_pop}
eucountries %>% 
  filter(sex_name == "both" & year_id == 2019) %>% 
  group_by(location_name, year_id) %>% 
  summarise(total_population = sum(total_population)) -> populations
populations$year_id <- as.integer(populations$year_id)
```

Draw ordered population bar diagram of the populations

```{r, eval = eval_pop}
options(scipen = 999)
ggplot(data = populations) +
  geom_bar(aes(x=reorder(location_name,total_population), y=total_population, group=year_id), position="dodge", stat="identity") +
  coord_flip() +
  labs(x="", y="population", title = "Countries total population in 2019")
```



```{r, eval = eval_pop, include=FALSE}
rm(countries)
rm(countries_population)
```



## IHME dataset

```{r, include = FALSE, eval=eval_IHME}
library(vroom)
library(fs)
library(purrr)
library(dplyr)
```


Exploring IHME dataset which is publicly available at [GBD webpage](http://ghdx.healthdata.org/gbd-results-tool?params=gbd-api-2019-public/706a4f69209d31dd557b6bc2b787a938). Dataset has prevalences and incidences of different disease by countries, age group and sex. You can download full dataset by this [link](http://ghdx.healthdata.org/gbd-results-tool?params=gbd-api-2019-permalink/e77f5cf5b1b5b77568d115f4878cd19b).


### Load few datasets

Firstly load libraries 

```{r, eval=eval_IHME}
library(vroom)
library(fs)
library(dplyr)
```

Downloaded data is stored in `data/ihmedata/`. List files under the data location by using `fs`-package and load few files

```{r, eval=eval_IHME}
files <- fs::dir_ls ("data/ihmedata/", recurse = T, glob = "*.csv")
vroom(file = files, n_max = 1000, id = "file_name")
```

Let's find out what two datasets has using command `spec()` and `summary()`

```{r, eval=eval_IHME, message=FALSE}
d1 <- vroom(file = files[1])
spec(d1)
summary(d1)
```

```{r, eval=eval_IHME}
d2 <- vroom(file = files[2])
spec(d2)
summary(d2)
```

Locations on each datasets

```{r, eval=eval_IHME}
length(unique(d1$location_name))
length(unique(d2$location_name))
```

Causes on each datasets

```{r, eval=eval_IHME}
length(unique(d1$cause_name))
length(unique(d2$cause_name))
```

```{r, eval=eval_IHME}
d1 %>% 
  group_by(location_name, cause_name) %>% 
  summarise(n = n(),
            val=sum(val)) %>% 
  arrange(n) %>% 
  tail(10)

d2 %>% 
  group_by(location_name, cause_name) %>% 
  summarise(n = n(),
            val=sum(val)) %>% 
  arrange(n) %>% 
  tail(10)
```


### Load all datasets

Load all datasets to one object and filter only Europe countries 

```{r, eval=eval_IHME, warning=FALSE}
europe <- c("Albania","Andorra","Armenia","Austria","Azerbaijan","Belarus","Belgium","Bosnia and Herzegovina","Bulgaria","Croatia","Cyprus","Czechia","Denmark","Estonia","Finland","France","Georgia","Germany","Greece","Hungary","Iceland","Ireland","Italy","Kazakhstan","Kosovo","Latvia","Liechtenstein","Lithuania","Luxembourg","Malta","Moldova","Monaco","Montenegro","Netherlands","North Macedonia","Norway","Poland","Portugal","Romania","Russian Federation","San Marino","Serbia","Slovakia","Slovenia","Spain","Sweden","Switzerland","Turkey","Ukraine","United Kingdom","Vatican City")

ihme <- NULL
for(i in seq_along(files)) {
  ihme <- ihme %>% 
    rbind(vroom(files[i], id = "file_name")) %>% 
    filter(location_name %in% europe)
}

```


### Summary 

Get basic info of the dataset

```{r, eval=eval_IHME}
summary(ihme)
```


```{r, eval=eval_IHME, include=FALSE}
rm(list=c("d1","d2"))
```

