
```{r setup_db, include=FALSE}
eval_loadflat=FALSE
eval_dbsave = FALSE
```

# Database

In this section we are going to clean and store datasets into `duckdb` database. We are going to filter datasets only to include these countries:

```{r, echo=FALSE}
europe <- c("Albania","Armenia","Austria","Azerbaijan","Belarus","Belgium","Bosnia and Herzegovina","Bulgaria","Croatia","Cyprus","Czechia","Denmark","Estonia","Finland","France","Georgia","Germany","Greece","Hungary","Iceland","Ireland","Italy","Kazakhstan","Latvia","Lithuania","Luxembourg","Malta","Republic of Moldova","Montenegro","Netherlands","North Macedonia","Norway","Poland","Portugal","Romania","Russian Federation","Serbia","Slovakia","Slovenia","Spain","Sweden","Switzerland","Turkey","Ukraine","United Kingdom")

europe
```

First load necessary packages

```{r}
library(fs)
library(dplyr)
library(vroom)
library(readxl)
```

## Load datasets

### Population dataset

Population dataset is in folder `data/IHME_GBD_2019_POP_2010_2019_SYA/`. Let's load all file locations

```{r, eval = eval_loadflat}
loc <- ("data/IHME_GBD_2019_POP_2010_2019_SYA/")
files <- dir_ls(loc, glob = "*.CSV")
files
```
Load all files and check available countries and years

```{r, eval = FALSE, include=FALSE}
POP <- NULL
for(i in seq_along(files)) {
  population <- vroom(files[i], id = "file_name")
  POP <- population %>% 
    group_by(location_name, year_id) %>% 
    summarise(total_population = sum(val, na.rm = T)) %>% 
    bind_rows(POP)
  rm(population)
}
unique(POP$year_id)
sort(as.character(unique(POP$location_name)))
```

Load all files and clean the datasets

```{r, eval = eval_loadflat, message=FALSE, warning=FALSE}
POP <- NULL
for(i in seq_along(files)) {
  population <- vroom(files[i], id = "file_name")
  
  POP <- population %>% 
    group_by(location_name, year_id, sex_name, age_group_name, age_group_id) %>% 
    summarise(total_population = sum(val, na.rm = T)) %>% 
    bind_rows(POP)
  
  rm(population)
}

## create a new age_group variable by re-adjusting age_group_name variable to integer
unique(POP$age_group_name)
POP$age_group <- POP$age_group_name
POP$age_group[POP$age_group == "<1 year"] <- 0
POP$age_group[POP$age_group == "95 plus"] <- 95
POP$age_group <- as.integer(POP$age_group)

## filter only europe countries
POP <- POP %>% 
  filter(location_name %in% europe) 

## Change location_names to universal
## change original to this: "Bosnia and Herz."> "Bosnia and Herzegovina"
## change original to this: Czechia" > ("Czech Rep.") > "Czech Republic"
## change original to this: "Russian Federation" > "Russia"
## change original to this: "North Macedonia" > "Macedonia"
## change original to this: "Republic of Moldova" > "Moldova" 
# unique(POP$location_name)
# POP[POP$location_name == "Bosnia and Herz.",] <- "Bosnia and Herzegovina"
POP$location_name[POP$location_name == "Czechia"] <- "Czech Republic"
POP$location_name[POP$location_name == "Russian Federation"] <- "Russia"
POP$location_name[POP$location_name == "North Macedonia"] <- "Macedonia"
POP$location_name[POP$location_name == "Republic of Moldova"] <- "Moldova"
```

Calculate Total population in 2019 of selected European countries

```{r, eval = eval_loadflat}
## Calculate Total population in 2019 of selected European countries
POP %>% 
  filter(sex_name == "both" & year_id == 2019) %>% 
  group_by(2019) %>%
  summarise(pop = sum(total_population)) -> total_population_2019
total_population_2019
```

### IHME dataset

Datasets are in folder `data/ihmedata/`. Load all the file location into a object.

```{r, eval = eval_loadflat}
files <- fs::dir_ls ("data/ihmedata/", recurse = T, glob = "*.csv")
# files
```

Load all files and filter only selected countries 

```{r, eval = eval_loadflat}
IHME <- NULL
for(i in seq_along(files)) {
  IHME <- IHME %>% 
    rbind(vroom(files[i], id = "file_name")) %>% 
    filter(location_name %in% europe)
}
```

```{r, eval = eval_loadflat, include=FALSE}
## Clean unnecessary objects
rm(list=c("files", "i", "loc"))
```

```{r, eval = eval_loadflat, include=FALSE}
## Check available countries and years
sort(as.character(unique(IHME$location_name)))
unique(IHME$year)
```

```{r, eval = eval_loadflat, include=FALSE}
## Filter dataset only to 
IHME <- IHME %>% 
    filter(location_name %in% europe)
```


```{r, eval = eval_loadflat, include=FALSE}
## Change location_names to universal
## change original to this: "Bosnia and Herz."> "Bosnia and Herzegovina"
## change original to this: Czechia" > ("Czech Rep.") > "Czech Republic"
## change original to this: "Russian Federation" > "Russia"
## change original to this: "North Macedonia" > "Macedonia"
## change original to this: "Republic of Moldova" > "Moldova" 
# sort(unique(IHME$location_name))
# IHME[IHME$location_name == "Bosnia and Herz.",] <- "Bosnia and Herzegovina"
IHME$location_name[IHME$location_name == "Czechia"] <- "Czech Republic"
IHME$location_name[IHME$location_name == "Russian Federation"] <- "Russia"
IHME$location_name[IHME$location_name == "North Macedonia"] <- "Macedonia"
IHME$location_name[IHME$location_name == "Republic of Moldova"] <- "Moldova"
```

### Calculator files

We have included a Excel file which holds data for the calculator. These data are taken from the article [Ameni et al. (2019) Cost-of-illness study of Obstructive Sleep Apnea Syndrome (OSAS) in Italy](https://cergas.unibocconi.eu/sites/default/files/files/Cost-of-illness-study-of-Obstructive-Sleep-Apnea-Syndrome-%2528OSAS%2529-in-Italy_Report%25281%2529.pdf)

```{r}
tbl1 <- readxl::read_xlsx("files/osa_tables.xlsx", sheet = "causes")
tbl4 <- readxl::read_xlsx("files/osa_tables.xlsx", sheet = "causes_simple")
tbl2 <- readxl::read_xlsx("files/osa_tables.xlsx", sheet = "osa_armeni")
tbl3 <- readxl::read_xlsx("files/osa_tables.xlsx", sheet = "osa_benjafield")
```


## Moving data to duckdb

In this section we move created datasets to duckdb database and trasform to *parquet* file. 

Further reading:

* <https://duckdb.org/docs/api/r>
* <https://duckdb.org/2021/12/03/duck-arrow.html>

Firstly install duckdb and arrow packages

```{r eval=FALSE}
install.packages("duckdb")
install.packages("arrow")
```

Create a connection and move datasets `POP` and `IHME` into duckdb. 

```{r, eval = eval_dbsave}
library(duckdb)
con = dbConnect(duckdb::duckdb(), dbdir="./data/sleep22.duckdb", read_only=FALSE)
dbWriteTable(con, "IHME", IHME, overwrite=TRUE)
dbWriteTable(con, "POP", POP, overwrite=TRUE)
dbWriteTable(con, "causes", tbl1, overwrite=TRUE) # old slapnea
dbWriteTable(con, "osa_armeni", tbl2, overwrite=TRUE)
dbWriteTable(con, "osa_benjafield", tbl3, overwrite=TRUE)
dbWriteTable(con, "causes_simple", tbl4, overwrite=TRUE)
dbDisconnect(conn = con)
```

### Transform to parquet

Transform whole database to parquet file system

```{r eval=eval_dbsave}
con = dbConnect(duckdb::duckdb(), dbdir="./data/sleep22.duckdb", read_only=FALSE)
if(dir.exists("data/parquet/")) fs::dir_delete("data/parquet/")
if(!dir.exists("data/parquet/")) dir.create("data/parquet/")
dbSendQuery(con, "EXPORT DATABASE 'data/parquet/' (FORMAT PARQUET);")
dbDisconnect(conn = con)
```

```{r}
#END
```


