```{r, include = FALSE}
eval_exploreihme <- TRUE #FALSE
# if(Sys.getenv("GLOBAL_EVAL") != "") eval_exploreihme <- Sys.getenv("GLOBAL_EVAL")
```


```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(vroom)
library(fs)
library(purrr)
library(dplyr)
```


# Explorative Data Analysis

In this section we are going to find out more about the data and it's variables, visualizing data and creating tables.

```{r}
library(duckdb)
library(arrow)
library(dplyr)


# ## Load from parquet files ----
# # Reads Parquet File to an Arrow Table
# arrow_slapnea <- arrow::read_parquet("data/parquet/3_slapnea.parquet", as_data_frame = FALSE)
# arrow_paf <- arrow::read_parquet("data/parquet/2_paf.parquet", as_data_frame = FALSE)
# arrow_ihme <- arrow::read_parquet("data/parquet/1_ihme.parquet", as_data_frame = FALSE)
# arrow_pop <- arrow::read_parquet("data/parquet/0_pop.parquet", as_data_frame = FALSE)
# 
# # Gets Database Connection
# con <- dbConnect(duckdb::duckdb())
# 
# # Registers arrow table as a DuckDB view
# arrow::to_duckdb(arrow_slapnea, table_name = "slapnea", con = con)
# arrow::to_duckdb(arrow_paf, table_name = "paf", con = con)
# arrow::to_duckdb(arrow_ihme, table_name = "ihme", con = con)
# arrow::to_duckdb(arrow_pop, table_name = "pop", con = con)
# 
# # dbListTables(con) # not working
# dbListFields(con, "ihme") # ihme list fields
# dbListFields(con, "pop") # pop list fields
# dbListFields(con, "paf") # paf list fields  
# dbListFields(con, "slapnea") # sleep apnea list fields


## Load duckdb database -----
con <- dbConnect(duckdb::duckdb(), dbdir="data/sleep22.duckdb", read_only=TRUE)
```

## IHME distinct values

Let's find out different distinct values of the variables

### measure_name

```{r}
dbGetQuery(con, "SELECT DISTINCT measure_name FROM ihme")
```

### sex_name

```{r}
dbGetQuery(con, "SELECT DISTINCT sex_name FROM ihme")
```

### age_name

```{r}
dbGetQuery(con, "SELECT DISTINCT age_name FROM ihme")
```

### metric_name

```{r}
dbGetQuery(con, "SELECT DISTINCT metric_name FROM ihme")
```

### year

```{r}
dbGetQuery(con, "SELECT DISTINCT year FROM ihme")
```

### cause_name

```{r}
dbGetQuery(con, "SELECT DISTINCT cause_id, cause_name FROM ihme ORDER BY cause_id")
```

### Calculator datas

```{r}
tbl(con, "causes") %>% 
  collect()
```
```{r}
tbl(con, "osa_armeni") %>% 
  collect()
```

```{r}
tbl(con, "osa_benjafield") %>% 
  collect()
```

## PLOT: Population in countries

```{r, fig.cap="Countries population in 2019."}
## First filter and group dataset to get total population by years and countries
tbl(con, "pop") %>% 
  filter(sex_name %in% c("both") & year_id == 2019) %>% 
  group_by(location_name, year_id) %>% 
  summarise(total_population = sum(total_population)) %>% 
  collect() -> dplot
dplot$year_id <- as.integer(dplot$year_id)

## Draw ordered population bar diagram of the populations
options(scipen = 999)
ggplot(data = dplot) +
  geom_bar(aes(x=reorder(location_name,total_population), y=total_population, group=year_id), position="dodge", stat="identity") +
  coord_flip() +
  labs(x="", y="population", title = "Countries total population in 2019")
```

## PLOT: Finland population 2019

Draw Finland population in 2019

```{r, eval = eval_exploreihme}
## filter dataset
tbl(con, "pop") %>% 
  filter(sex_name %in% c("female", "male") & location_name == "Finland" & year_id == 2019) %>% 
  collect() -> dplot

## plot age distribution
library(ggplot2)
ggplot(data = dplot) +
  geom_bar(aes(x=age_group, y=total_population, fill=sex_name), stat = "identity") + 
  labs(x="age", y="freq", title="Finland population 2019")
```



```{r, fig.cap="Countries population in 2019."}
## First filter and group dataset to get total population by years and countries
tbl(con, "pop") %>% 
  filter(sex_name %in% c("both") & year_id == 2019) %>% 
  group_by(location_name, year_id) %>% 
  summarise(total_population = sum(total_population)) %>% 
  collect() -> dplot
dplot$year_id <- as.integer(dplot$year_id)

## Draw ordered population bar diagram of the populations
options(scipen = 999)
ggplot(data = dplot) +
  geom_bar(aes(x=reorder(location_name,total_population), y=total_population, group=year_id), position="dodge", stat="identity") +
  coord_flip() +
  labs(x="", y="population", title = "Countries total population in 2019")
```




## TABLE: Prevalences

Create a table of all prevalences of disease available in the dataset from year 2019. 

```{r}
total_population <- tbl(con, "pop") %>% 
  filter(year_id == 2019 & sex_name == "both") %>% 
  summarise(pop=sum(total_population)) %>% 
  collect()

tbl(con, "ihme") %>% 
  filter(measure_name == "Prevalence" & sex_name == "Both" & metric_name == "Number") %>% 
  group_by(cause_id, cause_name) %>% 
  summarise(number = round(sum(val), 1)) %>% 
  collect() %>% 
  mutate(prevalence = round(100 * number / total_population$pop, 2)) %>% 
  arrange(cause_id) -> dplot_prev
```

```{r echo=FALSE}
DT::datatable(dplot_prev, rownames = FALSE)
```


## TABLE: Incidences

Create a table of all Incidence of disease available in the dataset from year 2019. 

```{r}
## Calculate Total  2019
tbl(con, "ihme")  %>% 
  filter(measure_name == "Incidence" & sex_name == "Both" & metric_name == "Number") %>% 
  group_by(cause_id, cause_name) %>% 
  summarise(number = round(sum(val), 1) #,
            # countries = length(unique(location_name))
            ) %>% 
  collect() %>% 
  mutate(percentage = round(100 * number / total_population$pop, 2)) %>% 
  arrange(cause_id) -> dplot_prev
```

```{r echo=FALSE}
DT::datatable(dplot_prev, rownames = FALSE)
```


## TABLE: Deaths

Create a table of all Incidence of disease available in the dataset from year 2019. 

```{r}
## Calculate Total  2019
tbl(con, "ihme")  %>% 
  filter(measure_name == "Deaths" & sex_name == "Both" & metric_name == "Number") %>% 
  group_by(cause_id, cause_name) %>% 
  summarise(number = round(sum(val), 1) #,
            # countries = length(unique(location_name))
            ) %>% 
  collect() %>% 
  mutate(percentage = round(100 * number / total_population$pop, 2)) %>% 
  arrange(cause_id) -> dplot_prev
```

```{r echo=FALSE}
DT::datatable(dplot_prev, rownames = FALSE)
```

## TABLE: Prevalences per countries

Group prevalences by countries and calculate percentage of the population (15-74 years old).

```{r}
filter_age_groups <- c("15 to 19", "20 to 24", "25 to 29", "30 to 34", "35 to 39", "40 to 44", "45 to 49", "50 to 54", "55 to 59", "60 to 64",
"65 to 69", "70 to 74")

## Prevalences of diseases by eucountries of the 15-74 years old population
tbl(con, "ihme")  %>% 
  filter(measure_name == "Prevalence" & sex_name == "Both" & age_name %in% filter_age_groups) %>% 
  filter(metric_name == "Number") %>% 
  group_by(location_name, cause_name) %>% 
  summarise(number = sum(val)) %>% 
  collect() %>% 
  mutate(location_name=as.factor(location_name)) -> dplot_prevalences

## filter population 
tbl(con, "pop") %>% 
  filter(sex_name == "both" & age_group > 14 & age_group < 75 & year_id == 2019) %>% 
  group_by(location_name) %>% 
  summarise(pop = sum(total_population)) %>% 
  collect() -> dplot_eucountries

## join tables
dplot_eucountries %>% 
  right_join(dplot_prevalences, by = "location_name") %>% 
  mutate(number=round(number,1),
         percentage=round(100*number/pop, 1)) -> dplot_prevalences
  
## Check the dimensions 
length(unique(dplot_prevalences$location_name)) ## locations
length(unique(dplot_prevalences$cause_name))    ## causes
```

```{r echo=FALSE}
DT::datatable(dplot_prevalences[, c("location_name", "cause_name", "number", "percentage")], filter = 'top', rownames = FALSE )
```

**Disconnect**

```{r}
dbDisconnect(con)
```

