```{r setup_visualize, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

## Load packages
library(duckdb)
library(arrow)
library(dplyr)
library(tidyr)
library(ggplot2)
library(leaflet)
library(hrbrthemes)
library(ggplot2)
library(tidyverse)
library(DT)

# hrbrthemes::import_roboto_condensed()

## We use same data as visualizer shiny app
slapnea22 <- readRDS(paste0(here::here(), "/data/slapnea22.RDS"))
europe_spdf <- readRDS(paste0(here::here(), "/data/europe_spdf.RDS"))
locations <- unique(slapnea22$location_name)
```

```{r}
## Load data

## DATA: sleep apnea costs rounded up
summary_slapnea_costs <- slapnea22 %>% 
  mutate(
    patient_direct_cost = round(patient_direct_cost, 0),
    patient_nonhealthcare_cost = round(patient_nonhealthcare_cost, 0),
    patient_productivity_cost = round(patient_productivity_cost, 0),
    patient_total_cost = round(patient_total_cost, 0)
  )

## DATA: PAtient direct cost per countries Create data for the barplot
slapnea22 %>% 
  select(location_name, patient_direct_cost, patient_nonhealthcare_cost, patient_productivity_cost) %>% 
  pivot_longer(c(patient_direct_cost, patient_nonhealthcare_cost, patient_productivity_cost)) -> d_patient_direct_cost

# unique(d_patient_direct_cost$name)

## Sleep apnea costs total MAX MEAN and MIN ----

## ROUNDED MAX OF EUROS PER PATIENT
slapnea22 %>% 
  select(location_name, patient_total_cost) %>% 
  group_by(location_name) %>% 
  summarise(sum = round(sum(patient_total_cost), -2)) %>% 
  arrange(sum) %>% 
  select(sum) %>% 
  tail(1) %>% 
  pull() -> max_value


slapnea22  %>% 
  select(location_name, patient_total_cost) %>% 
  group_by(location_name) %>% 
  summarise(sum = round(sum(patient_total_cost), -2)) %>% 
  arrange(sum) %>% 
  select(sum, location_name) %>% 
  tail(1) %>% 
  pull() -> max_location

## MEAN
slapnea22 %>% 
  select(location_name, patient_total_cost) %>% 
  summarise(mean = round(mean(patient_total_cost), -2)) %>% 
  arrange(mean) %>% 
  select(mean) %>% 
  tail(1) %>% 
  pull() -> mean_value
```



# Results without money correction

This section presents results of the non-money-corrected costs. Total costs are approximally `r mean_value` euros per patient. Maximum cost is `r max_value`  euros per patient in `r max_location`.

```{r plot_europe_cost1, message=FALSE}
## Europe map with cost per patient ----
## Thanks to <https://egallic.fr/en/european-map-using-r/>
library(ggplot2)
library(grid)
library(rworldmap)
library(mapproj)

# Get the world map
worldMap <- getMap()

## Check which are not included / names are different
tmp <- slapnea22 %>% group_by(location_name) %>% summarise(1) %>% select(location_name)
tmp <- tmp$location_name
# sort(as.character(worldMap$NAME))
tmp[!tmp %in% worldMap$NAME ]

## Need to change few location names
worldMap@data$NAME <- as.character(worldMap@data$NAME)
worldMap@data$NAME[worldMap@data$NAME == "Bosnia and Herz."] <- "Bosnia and Herzegovina" ## TODO check if names are correct in both datas
worldMap@data$NAME[worldMap@data$NAME == "Czech Rep."] <- "Czech Republic" ## TODO check if names are correct in both datas

## Check again
tmp[!tmp %in% worldMap$NAME ]
```


```{r plot_europe_cost2, message=FALSE}
# Select only the index of states
europea <- unique(summary_slapnea_costs$location_name)
indEU <- which(worldMap$NAME %in% tmp)

# View(worldMap$NAME)
# Extract longitude and latitude border's coordinates of members states
europeCoords <- lapply(indEU, function(i){
  df <- data.frame(worldMap@polygons[[i]]@Polygons[[1]]@coords)
  df$region =as.character(worldMap$NAME[i])
  colnames(df) <- list("long", "lat", "region")
  return(df)
})

europeCoords <- do.call("rbind", europeCoords)

# Add some data for each member
europeCoords$value <- summary_slapnea_costs$patient_total_cost[match(europeCoords$region, summary_slapnea_costs$location_name)] 

# Plot the map
ggplot() + 
  geom_polygon(data = europeCoords, 
               aes(x = long, y = lat, group = region, fill = value),
               colour = "black", size = 0.1) +
  coord_map(xlim = c(-22, 80),  ylim = c(30, 71)) +
  scale_fill_gradient(name = "Euros", high = "#FF0000FF", low = "#FFFF00FF", na.value = "grey50") +
  labs(title = "Sleep apnea cost in Europe", subtitle = "Total cost per patient",x="", y="") +
  hrbrthemes::theme_ipsum() +
  theme(
    panel.grid.minor = element_line(colour = NA),
    panel.background = element_rect(fill = NA, colour = NA),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(), axis.ticks.x = element_blank(),
    axis.ticks.y = element_blank(), axis.title = element_blank(),
    # rect = element_blank(),
    plot.margin = unit(0 * c(-1.5, -1.5, -1.5, -1.5), "lines")
    ) -> plot_europe_cost
ggsave("img/plot-europe-cost.png")
```


```{r}
# Plot the map
ggplot() + 
  geom_polygon(data = europeCoords, 
               aes(x = long, y = lat, group = region, fill = value),
               colour = "black", size = 0.1) +
  coord_map(xlim = c(-22, 80),  ylim = c(30, 71)) +
  scale_fill_gradient(name = "Euros", high = "#FF0000FF", low = "#FFFF00FF", na.value = "grey50") +
  labs(title = "", subtitle = "",x="", y="") +
  hrbrthemes::theme_ipsum() +
  theme(
    panel.grid.minor = element_line(colour = NA),
    panel.background = element_rect(fill = NA, colour = NA),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(), axis.ticks.x = element_blank(),
    axis.ticks.y = element_blank(), axis.title = element_blank(),
    # rect = element_blank(),
    plot.margin = unit(0 * c(-1.5, -1.5, -1.5, -1.5), "lines")
    ) -> plot_europe_cost
ggsave("img/plot-europe-cost-notext.png")
```


```{r, fig.cap="Estimated sleep apnea costs in Europe per patient in 2019."}
plot_europe_cost
```



```{r}
slapnea22 %>% 
  select(location_name, patient_total_cost, direct_cost) %>%
  # pivot_longer(c(patient_direct_cost, patient_nonhealthcare_cost, patient_productivity_cost)) %>% 
  collect() -> d_total_cost

ggplot(data = d_total_cost) +
  geom_text(aes(x=patient_total_cost, y=direct_cost, label=location_name), position=position_jitter(height=0.4) ) +
  hrbrthemes::theme_ipsum() +
  labs(title="Sleep apnea total cost", x="Euros / patient", y="Total") 

# ggplot(data = d_total_cost) +
#   geom_boxplot(aes(patient_total_cost)) +
#   # geom_text(aes(x=patient_total_cost, y=0, label=location_name), position=position_jitter(height=0.4) ) +
#   hrbrthemes::theme_ipsum() +
#   labs(title="Total cost variance", x="Euros / patient", y="") 
```

```{r, echo=FALSE, fig.width=10, fig.height=11, fig.fullwidth=TRUE}
## Barplot of the costs
## Create a bar plot by countries
ggplot(data = d_patient_direct_cost) +
  geom_bar(aes(x=reorder(location_name, value), y=value, fill=name), stat="identity") +
  coord_flip() +
  labs(x="", 
       y="euros", 
       fill="",
       title="Costs  by countries",
       subtitle = "Patient direct healthcare, direct non-healthcare and productivity lost costs") +
  hrbrthemes::theme_ipsum() +
  # scale_fill_discrete(labels=c('Direct healthcare cost', 'Direct non-helthcare cost', 'Productivity losses')) +
  scale_fill_brewer(palette = "Set2", labels=c('Direct healthcare cost', 'Direct non-helthcare cost', 'Productivity losses')) +
  scale_y_continuous(expand = c(0,0), limits = c(0, max_value*1.05)) +
  theme(plot.caption = element_text(hjust = 0, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot",
        legend.position = "right")  #NEW parameter

```


### Table: cost per patient

```{r}
DT::datatable(summary_slapnea_costs[, c("location_name", "patient_total_cost")] , rownames = FALSE)
```


## Direct costs

```{r, message=FALSE}
# Add some data for each member
europeCoords$value <- summary_slapnea_costs$patient_direct_cost[match(europeCoords$region, summary_slapnea_costs$location_name)] 

# Plot the map
ggplot() + 
  geom_polygon(data = europeCoords, 
               aes(x = long, y = lat, group = region, fill = value),
               colour = "black", size = 0.1) +
  coord_map(xlim = c(-22, 80),  ylim = c(30, 71)) +
  scale_fill_gradient(name = "Euros / patient", high = "#FF0000FF", low = "#FFFF00FF", na.value = "grey50") +
  labs(title = "Direct cost", subtitle="Sleep apnea cost in Europe", x="", y="") +
  hrbrthemes::theme_ipsum() +
  theme(
    # panel.grid.minor = element_line(colour = NA), 
    # panel.background = element_rect(fill = NA, colour = NA),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(), axis.ticks.x = element_blank(),
    axis.ticks.y = element_blank(), axis.title = element_blank(),
    # rect = element_blank(),
    plot.margin = unit(0 * c(-1.5, -1.5, -1.5, -1.5), "lines")
    ) -> plot_europe_direct_cost
ggsave("img/plot-europe-direct-cost.png")
```

```{r plot-europe-direct-cost, fig.align="center", fig.width=6, fig.height=6, fig.cap="Estimated sleep apnea direct cost per patient in Europe."}
plot_europe_direct_cost
```
### Table: direct cost per patient

```{r}
DT::datatable(summary_slapnea_costs[, c("location_name", "patient_direct_cost")] , rownames = FALSE)
```



## Direct non-healthcare costs

```{r plot-europe-directnonhealthcare-cost, message=FALSE}
# Add some data for each member
europeCoords$value <- summary_slapnea_costs$patient_nonhealthcare_cost[match(europeCoords$region, summary_slapnea_costs$location_name)] 

# Plot the map
ggplot() + 
  geom_polygon(data = europeCoords, 
               aes(x = long, y = lat, group = region, fill = value),
               colour = "black", size = 0.1) +
  coord_map(xlim = c(-22, 80),  ylim = c(30, 71)) +
  scale_fill_gradient(name = "Euros / patient", high = "#FF0000FF", low = "#FFFF00FF", na.value = "grey50") +
  labs(title = "Direct non-healthcare cost", subtitle="Sleep apnea cost in Europe", x="", y="") +
  hrbrthemes::theme_ipsum() +
  theme(
    # panel.grid.minor = element_line(colour = NA), 
    # panel.background = element_rect(fill = NA, colour = NA),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(), axis.ticks.x = element_blank(),
    axis.ticks.y = element_blank(), axis.title = element_blank(),
    # rect = element_blank(),
    plot.margin = unit(0 * c(-1.5, -1.5, -1.5, -1.5), "lines")
    ) -> plot_europe_directnonhealthcare_cost
ggsave("img/plot-europe-directnonhealthcare-cost.png")
```

```{r, fig.align="center", fig.width=6, fig.height=6, fig.cap="Estimated sleep apnea direct non-healthcare cost per patient in Europe."}
plot_europe_directnonhealthcare_cost
```

### Table: direct non-healthcare cost

```{r}
DT::datatable(summary_slapnea_costs[, c("location_name", "patient_nonhealthcare_cost")] , rownames = FALSE)
```

## Productivity loss costs

```{r plot-europe-productivityloss-cost, message=FALSE}
# Add some data for each member
europeCoords$value <- summary_slapnea_costs$patient_productivity_cost[match(europeCoords$region, summary_slapnea_costs$location_name)] 

# Plot the map
ggplot() + 
  geom_polygon(data = europeCoords, 
               aes(x = long, y = lat, group = region, fill = value),
               colour = "black", size = 0.1) +
  coord_map(xlim = c(-22, 80),  ylim = c(30, 71)) +
  scale_fill_gradient(name = "Euros / patient", high = "#FF0000FF", low = "#FFFF00FF", na.value = "grey50") +
  labs(title = "Productivity loss cost", subtitle="Sleep apnea cost in Europe", x="", y="") +
  hrbrthemes::theme_ipsum() +
  theme(
    # panel.grid.minor = element_line(colour = NA), 
    # panel.background = element_rect(fill = NA, colour = NA),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(), axis.ticks.x = element_blank(),
    axis.ticks.y = element_blank(), axis.title = element_blank(),
    # rect = element_blank(),
    plot.margin = unit(0 * c(-1.5, -1.5, -1.5, -1.5), "lines")
    ) -> plot_europe_productivityloss_cost
ggsave("img/plot-europe-productivityloss-cost.png")
```

```{r, fig.align="center", fig.width=6, fig.height=6, fig.cap="Estimated sleep apnea productivity losses cost per patient in Europe."}
plot_europe_productivityloss_cost
```

### Table: patient productivity cost

```{r}
DT::datatable(summary_slapnea_costs[, c("location_name", "patient_productivity_cost")] , rownames = FALSE)
```





