```{r setup_nexr, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# About

This project is systematically documented by using [bookdown](https://bookdown.org/) package. Bookdown is a package, which helps to write chapters and documents in R. We are also finding best practices on working with many/large csv-datasets by using packages such as [fs](https://www.tidyverse.org/blog/2018/01/fs-1.0.0/) and [vroom](https://www.tidyverse.org/blog/2019/05/vroom-1-0-0/), [tidyverse](https://www.tidyverse.org) and [duckdb](https://duckdb.org/).

```{r, include = FALSE}
options("width" = 82)
```

* [Memos](https://research.janimiettinen.fi/material/sleep22/index.html)
* [Github project](https://github.com/janikmiet/sleep22)


-----

# References

* [Bookdown package](https://bookdown.org/)

* [duckdb](https://duckdb.org/)

* [ghdx healthdata](http://ghdx.healthdata.org/)

* [Estimation source](https://cergas.unibocconi.eu/sites/default/files/files/Cost-of-illness-study-of-Obstructive-Sleep-Apnea-Syndrome-%2528OSAS%2529-in-Italy_Report%25281%2529.pdf)


**Index correction datas**

- <https://ec.europa.eu/eurostat/web/hicp/data/database>



# Testing output

## testing

```{r}
getwd()
```

```{r}
here::here()
```




