---
title: "Sleep Revolution: Sleep Apnea Cost"
author: "Jani Miettinen, Ida Alakörkkö, Anna But & Reijo Sund"
site: bookdown::bookdown_site
output:
  bookdown::gitbook:
    lib_dir: "book_assets"
    config:
      toc:
        collapse: section
        scroll_highlight: yes
        before: null
        after: null
      toolbar:
        position: fixed
      edit : null
      download: null
      search: yes
      fontsettings:
        theme: white
        family: sans
        size: 2
      sharing: no
  bookdown::pdf_book:
    keep_tex: yes
---

# Estimating costs of sleep apnea

Placeholder



<!--chapter:end:index.Rmd-->


# Data

Placeholder


## Population
### Load data into R
### Explore dataset
### Load and modify multiple files
### Included European countries
### Europe total population in 2019
### PLOT: Finland population 2019
### PLOT: European countries population
## IHME dataset
### Load few datasets
### Load all datasets
### Summary 

<!--chapter:end:01-flatfiles.Rmd-->


# Database

Placeholder


## Load datasets
### Population dataset
### IHME dataset
### Calculator files
## Moving data to duckdb
### Transform to parquet

<!--chapter:end:02-database.Rmd-->


# Explorative Data Analysis

Placeholder


## HME distinct values
### measure_name
### sex_name
### age_name
### metric_name
### year
### cause_name
### Calculator datas
## PLOT: Population in countries
## PLOT: Finland population 2019
## TABLE: Prevalences
## TABLE: Incidences
## TABLE: Deaths
## TABLE: Prevalences per countries

<!--chapter:end:03-eda.Rmd-->


# Calculations

Placeholder


## Prevalences from ihme
## Population from ihme
## Prevalences from ihme dataset
## Mortalities from ihme
## Combining with base values
## Table: prevalences and mortalities
## Save data to new database
## Costs
### Per Country
### Per Patient
### Add Money index to the calculation
### TABLE: Sleep Apnea costs
### Save dataset
## Validation
### Population
### Prevalences

<!--chapter:end:04-calculations.Rmd-->


# Results

Placeholder


### Table: cost per patient
## Direct costs
### Table: direct cost per patient
## Direct non-healthcare costs
### Table: direct non-healthcare cost
## Productivity loss costs
### Table: patient productivity cost
## Working on...
###  Leaflet map
### Another interactive example
### BASE MAP
### Confident intervals

<!--chapter:end:05-results.Rmd-->


# Money corrected results

Placeholder


### Table: total costs
## Direct costs
### Table: direct costs
## Direct non-healthcare costs
### Table: Direct non-healthcare costs
## Productivity loss costs
### Table: productivity loss cost

<!--chapter:end:06-results_corrected.Rmd-->


# Database for Calculator Shiny

Placeholder


## Population datasets
## Prevalences from ihme dataset
## Mortalities from ihme
## Add also base values
## Table: prevalences and mortalities
## Excel files
### OSA values from article
## Money index
## Save data to new database
### duckdb
### Parquet
## Data Based on Armeni OSA
## Data based on Benjafield OSA prevalence
## Map data
## Save and copy data

<!--chapter:end:07-newdb.Rmd-->

```{r setup_interactive, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Interactive applications

## Map visualization

```{r}
knitr::include_app(url = "https://janimiettinen.shinyapps.io/sleepapneacost/")
```

- URL: <https://janimiettinen.shinyapps.io/sleepapneacost/>

## Calculator

```{r}
knitr::include_app(url = "https://janimiettinen.shinyapps.io/sleepapneacalculator/")
```

- URL: <https://janimiettinen.shinyapps.io/sleepapneacalculator/>

<!--chapter:end:08-interactive.Rmd-->


# Side project

Placeholder


## Shiny App & todo ideas
## testing
## web page and shiny

<!--chapter:end:09-next.Rmd-->

```{r setup_article, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Estimating sleep apnea costs by open data science methods

Sleep apnea is disorder in which breathing repeatedly stops and stars during the sleep.


## Methods

There is no clear way how to calculate costs of sleep apnea. Patients can be undiscovered for long period time until they get sleep apnea diagnose.  Sleep apnea can cause ..... When patient is under diagnosed it means that patients can use healthcare services for many different reasons before they are diagnosed correctly and get help for sleep apnea.

One method to estimate obstructive sleep apnea cost is presented in article by Armeni et al. (2019) Cost-of-illness study of Obstructive Sleep Apnea Syndrome (OSAS) in Italy. Method estimates costs by top to down; diseases which are related to sleep apnea, are used for estimating costs which sleep apnea patients generates. Costs are divided intoo 3 groups; direct healthcare, direct noon healthcare and productivity lost costs. Direct costs includes hospitalizations, consultations, laboratory testing, drug or medical device consumption, etc. Direct non healthcare costs includes transportation costs and informal care (i.e. care provided by family). Productivity costs includes losses related to illness or death, patients and informal care time off work.


## Data

The Global Data Exchange (GHDx) data catalog provides health-related open data. The Institute for Metric Health and Evaluation (IHME) offers open data for for the catalog. In this project we collected data from 2019 for 42 different countries such as disease prevalences and population information.



## Building the estimation

Data sets are publicly available and downloadable from GHDx website. Data sets consists of the population and disease prevalances information. Data can be downloaded as (several) CSV-files to local computer. We decided to use `duckdb` database package for R to storage the data. `duckdb` creates local sql database, but for longer data storage we ended up saving datasets in parquet format, because in further `duckdb` version older databases might not work. Parquet datas can be pointed for `duckdb` and queries works as any sql databases.

After downloading population and prevalences datasets, let's store files into duckdb database for further analyses. First load necessary packages 

```{r}
library(fs)
library(dplyr)
library(vroom)
library(readxl)

selected_countries <- c("Albania","Armenia","Austria","Azerbaijan","Belarus","Belgium","Bosnia and Herzegovina","Bulgaria","Croatia","Cyprus","Czechia","Denmark","Estonia","Finland","France","Georgia","Germany","Greece","Hungary","Iceland","Ireland","Italy","Kazakhstan","Latvia","Lithuania","Luxembourg","Malta","Republic of Moldova","Montenegro","Netherlands","North Macedonia","Norway","Poland","Portugal","Romania","Russian Federation","Serbia","Slovakia","Slovenia","Spain","Sweden","Switzerland","Turkey","Ukraine","United Kingdom")
```

### Population dataset

Population dataset is in folder `data/IHME_GBD_2019_POP_2010_2019_SYA/`. Let's load all file locations

```{r, eval = eval_loadflat}
## List population datasets
loc <- ("data/IHME_GBD_2019_POP_2010_2019_SYA/")
files <- dir_ls(loc, glob = "*.CSV")

## Load all files and clean the dataset
POP <- NULL
for(i in seq_along(files)) {
  population <- vroom(files[i], id = "file_name")
  
  POP <- population %>% 
    group_by(location_name, year_id, sex_name, age_group_name, age_group_id) %>% 
    summarise(total_population = sum(val, na.rm = T)) %>% 
    bind_rows(POP)
  
  rm(population)
}

## create a new age_group variable by re-adjusting age_group_name variable to integer
unique(POP$age_group_name)
POP$age_group <- POP$age_group_name
POP$age_group[POP$age_group == "<1 year"] <- 0
POP$age_group[POP$age_group == "95 plus"] <- 95
POP$age_group <- as.integer(POP$age_group)

## filter only selected_countries 
POP <- POP %>% 
  filter(location_name %in% selected_countries) 

## Change location_names to universal
POP$location_name[POP$location_name == "Czechia"] <- "Czech Republic"
POP$location_name[POP$location_name == "Russian Federation"] <- "Russia"
POP$location_name[POP$location_name == "North Macedonia"] <- "Macedonia"
POP$location_name[POP$location_name == "Republic of Moldova"] <- "Moldova"

## Calculate Total population in 2019 of selected selected_countries
POP %>% 
  filter(sex_name == "both" & year_id == 2019) %>% 
  group_by(2019) %>%
  summarise(pop = sum(total_population)) -> total_population_2019
total_population_2019
```

### IHME dataset

Datasets are in folder `data/ihmedata/`. Load all the file location into a object.

```{r, eval = eval_loadflat}
## List IHME datasets
files <- fs::dir_ls ("data/ihmedata/", recurse = T, glob = "*.csv")

## Load all files and filter only selected_countries
IHME <- NULL
for(i in seq_along(files)) {
  IHME <- IHME %>% 
    rbind(vroom(files[i], id = "file_name")) %>% 
    filter(location_name %in% selected_countries)
}

## Filter dataset only to specific countries
IHME <- IHME %>% 
    filter(location_name %in% selected_countries)

## Change location_names to universal
IHME$location_name[IHME$location_name == "Czechia"] <- "Czech Republic"
IHME$location_name[IHME$location_name == "Russian Federation"] <- "Russia"
IHME$location_name[IHME$location_name == "North Macedonia"] <- "Macedonia"
IHME$location_name[IHME$location_name == "Republic of Moldova"] <- "Moldova"
```

### Calculator files

We have included a Excel file which holds data for the calculator. These data are taken from the article [Ameni et al. (2019) Cost-of-illness study of Obstructive Sleep Apnea Syndrome (OSAS) in Italy](https://cergas.unibocconi.eu/sites/default/files/files/Cost-of-illness-study-of-Obstructive-Sleep-Apnea-Syndrome-%2528OSAS%2529-in-Italy_Report%25281%2529.pdf)

```{r}
tbl1 <- readxl::read_xlsx("data/osa_tables.xlsx", sheet = "causes")
tbl2 <- readxl::read_xlsx("data/osa_tables.xlsx", sheet = "osa_armeni")
```


### Storing the data to duckdb

In this section we move created datasets to duckdb database and trasform to *parquet* file.  Firstly install duckdb and arrow packages

```{r eval=FALSE}
install.packages("duckdb")
install.packages("arrow")
```

Create a connection and move datasets `POP` and `IHME` into duckdb. 

```{r, eval = eval_dbsave}
library(duckdb)
con = dbConnect(duckdb::duckdb(), dbdir="./data/sleep22.duckdb", read_only=FALSE)
dbWriteTable(con, "IHME", IHME, overwrite=TRUE)
dbWriteTable(con, "POP", POP, overwrite=TRUE)
dbWriteTable(con, "slapnea", tbl1, overwrite=TRUE)
dbWriteTable(con, "paf", tbl2, overwrite=TRUE)
dbDisconnect(conn = con)
```

Transform whole database to parquet file system

```{r eval=eval_dbsave}
## Transform to parquet
con = dbConnect(duckdb::duckdb(), dbdir="./data/sleep22.duckdb", read_only=FALSE)
if(dir.exists("data/parquet/")) fs::dir_delete("data/parquet/")
if(!dir.exists("data/parquet/")) dir.create("data/parquet/")
dbSendQuery(con, "EXPORT DATABASE 'data/parquet/' (FORMAT PARQUET);")
dbDisconnect(conn = con)
```


## Database for Calculator Shiny

In this chapter we are going to use same methods as in [calculation chapter](), but we are going to create database, which we can use in Shiny Application, where user can add his input values for sleep apnea prevalences, condition prevalences and costs.


```{r}
library(duckdb)
library(arrow)
library(dplyr)
options(scipen = 999)

## Load duckdb
con = dbConnect(duckdb::duckdb(), dbdir="data/sleep22.duckdb", read_only=TRUE)

```

## Population datasets

Calculating 15-74 years old populations

```{r}
## Both, female and male population (15-74yrs) in wide format
library(tidyr)
tbl(con, "pop") %>% 
  filter(age_group > 14 & age_group < 75 & year_id == 2019) %>% 
  group_by(location_name, sex_name) %>% 
  summarise(pop = sum(total_population)) %>% 
  collect() %>% 
  pivot_wider(names_from = sex_name, values_from = pop) %>% 
  rename(pop_1574_both = both,
         pop_1574_female = female,
         pop_1574_male = male) -> population_1574
head(population_1574)
```

Calculating 30 to 69 years old population

```{r}
## Both, female and male population (30-69yrs) in wide format
library(tidyr)
tbl(con, "pop") %>% 
  filter(age_group > 29 & age_group < 70 & year_id == 2019) %>% 
  group_by(location_name, sex_name) %>% 
  summarise(pop = sum(total_population)) %>% 
  collect() %>% 
  pivot_wider(names_from = sex_name, values_from = pop) %>% 
  rename(pop_3069_both = both,
         pop_3069_female = female,
         pop_3069_male = male) -> population_3069
head(population_3069)
```

```{r}
# Join two populations to one dataset
pop <- population_1574 %>% 
  left_join(population_3069, by = "location_name")
```

Total populations

```{r}
## Both, female and male population (15-74yrs) in wide format
library(tidyr)
tbl(con, "pop") %>% 
  filter(year_id == 2019) %>% 
  group_by(location_name, sex_name) %>% 
  summarise(pop = sum(total_population)) %>% 
  collect() %>% 
  pivot_wider(names_from = sex_name, values_from = pop) %>% 
  rename(pop_both = both,
         pop_female = female,
         pop_male = male) -> popu_info
head(popu_info)
```


```{r}
rm(list=c("population_1574", "population_3069"))
```



## Prevalences from ihme dataset

Calculating prevalences percentages by joining prevalences and population information

```{r}
## Get cause_id's which we can use from IHME dataset
tbl(con, "causes") %>% 
  filter(type == "prevalence") %>%
  select(cause_id) %>% 
  collect() -> causes
causes <- causes$cause_id[!is.na(causes$cause_id)]

## Get number of prevalences per location and causes from IHME (age: 15-74 and sex: both)
tbl(con, "ihme") %>% 
  filter(cause_id %in% causes & metric_name == "Number" & age_id > 7 & age_id < 20 & sex_id == 3) %>% 
  group_by(location_name, cause_id, cause_name) %>% 
  summarise(prevalence_number = sum(val)) %>% 
  arrange(location_name, cause_id) %>% 
  collect() -> prevalences_ihme
## Calculate Prevalence Rates: join tables prevalences_ihme & pop
prevalences_ihme %>% 
  right_join(pop, by = "location_name") %>% 
  mutate(prevalence = prevalence_number / pop_1574_both,
         age_group = "1574")  %>% 
  select(location_name, cause_id, cause_name, prevalence, age_group) -> prevalences_ihme


## Prevalences data for 3069 ages
## Get number of prevalences per location and causes from IHME (age: 30-69 and sex: both)
tbl(con, "ihme") %>%
  filter(cause_id %in% causes & metric_name == "Number" & age_id > 10 & age_id < 19 & sex_id == 3) %>%
  group_by(location_name, cause_id, cause_name) %>%
  summarise(prevalence_number = sum(val)) %>%
  arrange(location_name, cause_id) %>%
  collect() -> prevalences_ihme2
## Calculate Prevalence Rates: join tables prevalences_ihme & pop
prevalences_ihme2 %>%
  right_join(pop, by = "location_name") %>%
  mutate(prevalence = prevalence_number / pop_3069_both,
         age_group = "3069")  %>%
  select(location_name, cause_id, cause_name, prevalence, age_group) -> prevalences_ihme2


## Join these two
prevalences_ihme %>%
  rbind(prevalences_ihme2) -> prevalences_ihme

head(prevalences_ihme)
```



## Mortalities from ihme

Calculating mortalities for two diseases and calculating percentages by joining mortality and population information

- 1023 	Other cardiovascular and circulatory diseases
- 294  	All causes

```{r}
## Get cause_id's which base values we can use to overwrite from IHME data
tbl(con, "causes") %>% 
  filter(type == "mortality") %>% 
  select(cause_id) %>% 
  collect() -> causes
causes <- causes$cause_id[!is.na(causes$cause_id)]

## Get mortality values from ihme
tbl(con, "ihme") %>% 
  filter(cause_id %in% causes & measure_name == "Deaths" & metric_name == "Number" & age_id > 7 & age_id < 20 & sex_id == 3) %>% 
  group_by(location_name, cause_id, cause_name) %>% 
  summarise(mortality_number = sum(val)) %>% 
  arrange(location_name, cause_id) %>% 
  collect() -> mortalities_ihme
## Calculate Rates: join tables mortalities_ihme & pop
mortalities_ihme %>% 
  right_join(pop, by = "location_name") %>% 
  mutate(mortality = mortality_number / pop_1574_both,
         age_group = "1574") %>% 
  select(location_name, cause_id, cause_name, mortality, age_group) -> mortalities_ihme


## Mortalities data for 3069 ages
## Get mortality values from ihme (30 - 69yrs)
tbl(con, "ihme") %>%
  filter(cause_id %in% causes & measure_name == "Deaths" & metric_name == "Number" & age_id > 10 & age_id < 19 & sex_id == 3) %>%
  group_by(location_name, cause_id, cause_name) %>%
  summarise(mortality_number = sum(val)) %>%
  arrange(location_name, cause_id) %>%
  collect() -> mortalities_ihme2
## Calculate Rates: join tables mortalities_ihme & pop
mortalities_ihme2 %>%
  right_join(pop, by = "location_name") %>%
  mutate(mortality = mortality_number / pop_3069_both,
         age_group = "3069") %>%
  select(location_name, cause_id, cause_name, mortality, age_group) -> mortalities_ihme2



## Join these two
mortalities_ihme %>%
  rbind(mortalities_ihme2) -> mortalities_ihme

head(mortalities_ihme)
```

```{r}
## Join mortalities and prevalance data. Add name to describe value and rename 'mortality'/'prevalence' to 'value'.
prevalences_ihme$type <- "prevalence"
mortalities_ihme$type <- "mortality"
ihme <- prevalences_ihme %>%
  rbind(mortalities_ihme) %>% 
  mutate(ihme = ifelse(is.na(prevalence), mortality, prevalence)) %>% 
  select(location_name, cause_id, cause_name, type, ihme, age_group) %>% 
  arrange(location_name, type, cause_id)
head(ihme)
```


```{r}
rm(list=c("mortalities_ihme", "prevalences_ihme", "mortalities_ihme2", "prevalences_ihme2"))
```



## Add also base values

Base data (from excel) and country specific prevalences & mortalities


```{r}
## Base data. We need paf info for every country
prevalences_base <- NULL
for(cntry in unique(ihme$location_name)){
  # print(cntry)
  # tbl(con, "causes") %>% 
  tbl(con, "slapnea") %>% 
    collect() %>% 
    mutate(location_name = cntry,
           condition = Conditon) %>% 
    select("location_name", 
           "cause_id", 
           "condition", 
           "OR",
           "RR",
           "PAF", 
           "prevalence_base_italy", 
           "annual_direct_healthcare_cost",
           "annual_direct_nonhealthcare_cost", 
           "annual_productivity_losses_cost") %>% 
    rbind(prevalences_base) -> prevalences_base
}
head(prevalences_base) #1296
# 1296 *2 2592
```

With base values, join new calculated prevalences/mortalities from IHME

```{r}
## Join base data and new calculated prevalences, age_group 1574
prevalences1 <- prevalences_base %>%
  left_join(ihme %>% filter(age_group == "1574") , by = c("location_name", "cause_id")) %>% 
  mutate(age_group = "1574") %>%
  select(location_name, age_group, condition, cause_id, cause_name, ihme, prevalence_base_italy, OR, RR, PAF, annual_direct_healthcare_cost, annual_direct_nonhealthcare_cost, annual_productivity_losses_cost) %>% 
  right_join(
    pop %>% 
      select(location_name, pop_1574_both, pop_1574_female, pop_1574_male) %>% 
      rename(pop_both=pop_1574_both, pop_female=pop_1574_female, pop_male=pop_1574_male),
    by = "location_name") ## Add the population to all rows

## Join base data and new calculated prevalences, age_group 3069
prevalences2 <- prevalences_base %>%
  left_join(ihme %>% filter(age_group == "3069") , by = c("location_name", "cause_id")) %>% 
  mutate(age_group = "3069") %>%
  select(location_name, age_group, condition, cause_id, cause_name, ihme, prevalence_base_italy, OR, RR, PAF, annual_direct_healthcare_cost, annual_direct_nonhealthcare_cost, annual_productivity_losses_cost) %>% 
  right_join(
     pop %>% 
      select(location_name, pop_3069_both, pop_3069_female, pop_3069_male) %>% 
      rename(pop_both=pop_3069_both, pop_female=pop_3069_female, pop_male=pop_3069_male),
    by = "location_name") ## Add the population to all rows

# Join age_group prevalence  tables
prevalences <- prevalences1 %>% 
  rbind(prevalences2)

head(prevalences)

rm(list = c("prevalences1", "prevalences2", "prevalences_base", "ihme"))
```




## Table: prevalences and mortalities

```{r}
DT::datatable(prevalences, rownames = FALSE)
```

## Excel files

### OSA values from article

```{r}
### TODO left here, get prevelance data for slider
library(data.table)
## Read from source excel
osa <- readxl::read_xlsx("data/osa_tables.xlsx", sheet = "osa_armeni")
## Sleep apnea prevalences
osanew <- readxl::read_xlsx("data/osa_tables.xlsx", sheet = "osa_benjafield")
osanew$country
## How many countries are needed
length(unique(prevalences$location_name))
## Check which countries are not included in excel
length(osanew$country[osanew$country %in% unique(prevalences$location_name)])
## 3 countries missing, check names which are missing
unique(prevalences$location_name)[!unique(prevalences$location_name) %in% osanew$country]
## these are not included # [1] "Andorra"        "Monaco"         "San Marino"  

## Filter dataset and Calculate Mild, Moderate, Severe values separately
osanew <-osanew %>% 
  filter(country %in% unique(prevalences$location_name)) %>% 
  mutate(
    Mild = `Mild-Moderate-Severe` - `Moderate-Severe`,
    `Male_Moderate-Severe` = 1.3823 * `Moderate-Severe`, # Gender proportion is calculated from Italy data from Armeni article data
    `Female_Moderate-Severe` = 0.6508 * `Moderate-Severe`, # Gender proportion is calculated from Italy data from Armeni article data
    `Both_Moderate-Severe` = `Moderate-Severe`, # Gender proportion is calculated from Italy data from Armeni article data
    Both_Mild = Mild,
    Male_Mild =  0.9519 * Mild, # Gender proportion is calculated from Italy data from Armeni article data
    Female_Mild = 1.0440 * Mild # Gender proportion is calculated from Italy data from Armeni article data
  )

osanew <- osanew %>% 
  mutate(
    ## TODO check these with comments
    Female_Moderate = 0.5342 * `Female_Moderate-Severe`, # Moderate proportion percentage is calculated from Italy data from Armeni article data
    Female_Severe =  0.4658 * `Female_Moderate-Severe`,   # Severe proportion percentage is calculated from Italy data from Armeni article data
    Male_Moderate = 0.4004 * `Male_Moderate-Severe`, # Moderate proportion percentage is calculated from Italy data from Armeni article data
    Male_Severe =  0.5996 * `Male_Moderate-Severe` ,
    Both_Moderate = 0.445907 * `Both_Moderate-Severe`,
    Both_Severe = 0.554093 * `Both_Moderate-Severe`
  )

## To longer format
library(tidyr)
osanew <- osanew %>%
  tidyr::pivot_longer(c(`Male_Moderate-Severe`, `Male_Moderate`, `Male_Severe`, Male_Mild, `Female_Moderate-Severe`, `Female_Moderate`, `Female_Severe`,  Female_Mild, `Both_Moderate-Severe`, `Both_Moderate`, `Both_Severe`, Both_Mild), names_to = "var") %>%
  rename(rate=value,
         location_name = country) %>%
  mutate(
    ## Get gender and var right
    gender = sub("_.*", "", var),
    var = sub('.+_(.+)', '\\1', var) 
    ) %>% 
  select(location_name, var, gender, rate)
```

## Money index

```{r}
## First try HICP index, what countries are included -----
## https://appsso.eurostat.ec.europa.eu/nui/show.do?dataset=prc_hicp_aind&lang=en
hicp <- readxl::read_xls("data/prc_hicp_aind.xls")
hicp$`GEO/TIME`[hicp$`GEO/TIME` == "Germany (until 1990 former territory of the FRG)"] <- "Germany" ## change name of the country

## Calculate correction index to all other countries
hicp %>% 
  filter(`GEO/TIME` %in% unique(prevalences$location_name)) %>% 
  rename(location_name = `GEO/TIME`) %>% 
  select(location_name, `2019`) %>% 
  mutate(index = `2019` / hicp$`2019`[hicp$`GEO/TIME` == "Italy"],
         ver = "EuroStat") %>% 
  select(location_name, index, ver) -> money_correction

## This is missing few countries so their index correction is 1 (equal to Italy)
## TODO find a way to correct these
# unique(prevalences_base$location_name[!prevalences_base$location_name %in% money_correction$location_name])
```


```{r}
dbDisconnect(con, shutdown=TRUE)
```


## Save data to new database

### duckdb

```{r, eval=eval_prevalences_savedb}
# library(duckdb)
# con = dbConnect(duckdb::duckdb(), dbdir="./data/sleep22_shiny.duckdb", read_only=FALSE)
# dbWriteTable(con, "prevalences", prevalences, overwrite=TRUE)
# dbWriteTable(con, "pop", pop, overwrite=TRUE)
# dbWriteTable(con, "popu_info", popu_info, overwrite=TRUE)
# dbWriteTable(con, "osa", osa, overwrite=TRUE)
# dbWriteTable(con, "osanew", osanew, overwrite=TRUE)
# dbWriteTable(con, "money_correction", money_correction, overwrite=TRUE)
# dbDisconnect(con)
# 
# con = dbConnect(duckdb::duckdb(), dbdir="./data/sleep22_shiny.duckdb", read_only=FALSE)
# if(dir.exists("data/parquet_shiny/")) fs::dir_delete("data/parquet_shiny/")
# if(!dir.exists("data/parquet_shiny/")) dir.create("data/parquet_shiny/")
# dbSendQuery(con, "EXPORT DATABASE 'data/parquet_shiny/' (FORMAT PARQUET);") # copy komento? COPY taulu TO ... FORMAT 

# dbDisconnect(conn = con)
```

```{r, eval=FALSE}
## Copy newdb to sleep22calculator
# fils <- c(paste0(here::here(), "/data/sleep22_shiny.duckdb"), paste0(here::here(), "/data/sleep22_shiny.duckdb.wal"))
# fs::file_copy(fils, "/Users/japmiett/projects/sleep22calculator/", overwrite = T)
```

### Parquet



```{r}
## Testing another way to save database only to parquet files
con = dbConnect(duckdb::duckdb(), dbdir=":MEMORY:", read_only=FALSE)
## Register tbls
duckdb::duckdb_register(conn = con, name = "prevalences", df = prevalences)
duckdb::duckdb_register(conn = con, name = "pop", df = pop)
duckdb::duckdb_register(conn = con, name = "popu_info", df = popu_info)
duckdb::duckdb_register(conn = con, name = "osa", df = osa)
duckdb::duckdb_register(conn = con, name = "osanew", df = osanew)
duckdb::duckdb_register(conn = con, name = "money_correction", df = money_correction)
## Save to parquet
dbSendQuery(con, "COPY (SELECT * FROM prevalences) TO 'data/parquet_shiny/prevalences.parquet' (FORMAT 'parquet');")
dbSendQuery(con, "COPY (SELECT * FROM pop) TO 'data/parquet_shiny/pop.parquet' (FORMAT 'parquet');")
dbSendQuery(con, "COPY (SELECT * FROM popu_info) TO 'data/parquet_shiny/popu_info.parquet' (FORMAT 'parquet');")
dbSendQuery(con, "COPY (SELECT * FROM osa) TO 'data/parquet_shiny/osa.parquet' (FORMAT 'parquet');")
dbSendQuery(con, "COPY (SELECT * FROM osanew) TO 'data/parquet_shiny/osanew.parquet' (FORMAT 'parquet');")
dbSendQuery(con, "COPY (SELECT * FROM money_correction) TO 'data/parquet_shiny/money_correction.parquet' (FORMAT 'parquet');")
dbDisconnect(conn = con)
```

```{r}
## Copy to sleep22calculator
fils <- list.files("data/parquet_shiny/", full.names = T) 
fs::file_copy(fils, "/Users/japmiett/projects/sleep22calculator/data/", overwrite = T)
```







# Data for Visualizer Shiny

We need this function for calculating PAF from Odd Ratio. 

```{r}
## PAF ODD RATIO formula -----
## Give only decimals as parameters
# PD  ## having a disease, prevalence
# PE  ## exposed, sleep apnea prevalence?
# PE_ ##  unexposed, 
paf_or <- function(OR, PD, PE){
  PD = PD * 100
  PE = PE * 100
  PE_ = 100 - PE
  VALUE1 = (PD * (1 - OR) + PE_ + OR * PE + sqrt( (PD * (1 - OR) + PE_ + OR * PE )^2 - 4 * PE_ * (1 - OR) *PD )) / (2 * PE_ * (1 - OR))
  VALUE2 = (PD * (1 - OR) + PE_ + OR * PE - sqrt( (PD * (1 - OR) + PE_ + OR * PE )^2 - 4 * PE_ * (1 - OR) *PD )) / (2 * PE_ * (1 - OR))
  VALUE <- ifelse(VALUE1 < 100 & VALUE1 > 0, VALUE1, VALUE2)
  PAF = 1 - ((100 * VALUE) / PD)
  return(PAF)
}

```


## Data Based on Armeni OSA

```{r, eval = FALSE}
## We need 4 precalculate datasets/values
## Data for Armeni OSA table and no money index correction
## Data for Armeni OSA table with money index correction
## Data for Benjafield OSA table and no money index correction
## Data for Benjafield OSA table with money index correction

## 1) Data with Armeni OSA table and no money index correction
prevalences %>% 
  filter(age_group == "1574") %>% ## TODO Armeni / Benjafield
  group_by(location_name) %>%
  mutate(
    # Select prevalence which to use
    prevalence = ifelse(is.na(ihme), prevalence_base_italy, ihme), 
    ## PAF calculation for Risk Ratio or Odds Ratio:
    # osa_value = osa$rate[osa$gender == "Both" & osa$var == "Moderate-Severe"], ## TODO this needs to be specific  Armeni / Benjafield
    osa_value = (osa$rate[osa$gender == "Male" & osa$var == "Moderate-Severe"] * pop_male + osa$rate[osa$gender == "Female" & osa$var == "Moderate-Severe"] * pop_female) / pop_both,
    PAFRR = ifelse(!is.na(RR), (osa_value * (RR - 1) / (osa_value * (RR - 1) + 1)), NA), 
    PAFOR = ifelse(!is.na(OR), paf_or(OR, prevalence, osa_value), NA),
    PAF = ifelse(is.na(PAFOR), PAFRR, PAFOR),
    ## Prevalents per conditions
    prevalent_cases = prevalence * pop_both, ## Taudin prevalenssi * populaatio, ok
    prevalent_cases_influenced_osa = PAF * prevalent_cases, ## PAF * prevalent_cases, ok
    ## Costs per conditions
    direct_cost = prevalent_cases_influenced_osa * annual_direct_healthcare_cost, ## ok
    direct_non_healthcare_cost = prevalent_cases_influenced_osa * annual_direct_nonhealthcare_cost, ## ok
    productivity_lost_cost = prevalent_cases_influenced_osa * annual_productivity_losses_cost, ## ok
    direct_cost = ifelse(is.na(direct_cost), 0 , direct_cost),
    direct_non_healthcare_cost = ifelse(is.na(direct_non_healthcare_cost), 0 , direct_non_healthcare_cost),
    productivity_lost_cost = ifelse(is.na(productivity_lost_cost), 0 , productivity_lost_cost),
    total_costs = direct_cost + direct_non_healthcare_cost + productivity_lost_cost) -> slapnea_cost1
## Cpst per patient and sums
slapnea_cost1 %>% 
  group_by(location_name, pop_female, pop_male) %>% 
  summarise(
    ## Sums of costs
    direct_cost = sum(direct_cost, na.rm = T),
    direct_non_healthcare_cost = sum(direct_non_healthcare_cost, na.rm = T),
    productivity_lost_cost = sum(productivity_lost_cost, na.rm = T)
    )  %>% 
  mutate(
    ## Absolute values for dividing the cost per patient
    absolute_value_severe_moderate = ( (pop_female * osa$rate[osa$var == "Moderate" & osa$gender == "Female"]) + (pop_female * osa$rate[osa$var == "Severe" & osa$gender == "Female"]) + (pop_male * osa$rate[osa$var == "Moderate" & osa$gender == "Male"]) + (pop_male * osa$rate[osa$var == "Severe" & osa$gender == "Male"])), ## ok
    # absolute_value_mild = ( pop_female * osa$rate[osa$var == "Mild" & osa$gender == "Female"]) + (pop_male * osa$rate[osa$var == "Mild" & osa$gender == "Male"] ), ## ok
    ## Costs per patients
    patient_direct_cost = direct_cost / absolute_value_severe_moderate, ## ok
    patient_nonhealthcare_cost = direct_non_healthcare_cost / absolute_value_severe_moderate, ## ok
    patient_productivity_cost = productivity_lost_cost / absolute_value_severe_moderate, ## ok
    ## Total cost per patient
    patient_total_cost = patient_direct_cost + patient_nonhealthcare_cost + patient_productivity_cost ## ok
  ) -> slapnea22
## Write to calculation database
saveRDS(slapnea22, file = "data/slapnea22.RDS")

# View(slapnea_cost2[, c("location_name", "patient_total_cost")])
# Results: costs are almost right, 50 - 100 difference. WHYYYY???

slapnea22 %>% 
  left_join(money_correction, by = "location_name") %>% 
  mutate(
    index= ifelse(is.na(index), 1, index),
    direct_cost = direct_cost * index,
    direct_non_healthcare_cost = direct_non_healthcare_cost * index,
    productivity_lost_cost = productivity_lost_cost * index,
    total_cost = direct_cost + direct_non_healthcare_cost +productivity_lost_cost,
    ## Per patient
    patient_direct_cost = direct_cost / absolute_value_severe_moderate, ## ok
    patient_nonhealthcare_cost = direct_non_healthcare_cost / absolute_value_severe_moderate, ## ok
    patient_productivity_cost = productivity_lost_cost / absolute_value_severe_moderate, ## ok
    patient_total_cost = patient_direct_cost + patient_nonhealthcare_cost + patient_productivity_cost ## ok
  ) -> slapnea22_eurostat
## Write to calculation database
saveRDS(slapnea22_eurostat, file = "data/slapnea22_eurostat.RDS")


# ## Make money correction
# if(input$money_index == "EuroStat '19"){
#   dplot <- dplot %>%
#     left_join(money_correction, by = "location_name") %>%
#     mutate(corrected = ifelse(is.na(index), FALSE, TRUE),
#            index = ifelse(is.na(index), 1, index),
#            direct_cost = direct_cost * index,
#            direct_non_healthcare_cost = direct_non_healthcare_cost * index,
#            productivity_lost_cost = productivity_lost_cost * index,
#            total_costs = total_costs * index)
# }

```

## Data based on Benjafield OSA prevalence

```{r}
## Data for Benjafield OSA table and no money index correction
## Data for Benjafield OSA table with money index correction

## Missä kohtaa laskennassa käytetään uniapnean prevalenssia?

## 1) Data with Armeni OSA table and no money index correction
## Delete   "San Marino" "Monaco"     "Andorra"
prevalences <- prevalences %>% 
  filter(!location_name %in% c("San Marino", "Monaco", "Andorra"))

prevalences %>% 
  filter(age_group == "3069") %>% ##  Benjafield POP
  group_by(location_name) %>%
  mutate(
    # Select prevalence which to use
    prevalence = ifelse(is.na(ihme), prevalence_base_italy, ihme), 
    ## PAF calculation for Risk Ratio or Odds Ratio:
    ## TODO how  to make this connection in tidyverse?????
    ## "San Marino" "Monaco"     "Andorra"  are missing from osanew. This needs to be fixed also in shinyapp!
    ## Add these manually, italy values.
    # osa_value = osa$rate[osa$gender == "Both" & osa$var == "Moderate-Severe"], ## TODO this needs to be specific  Armeni / Benjafield
    osa_value = (osanew$rate[osanew$gender == "Male" & osanew$var == "Moderate-Severe" & osanew$location_name == location_name] * pop_male + osanew$rate[osanew$gender == "Female" & osanew$var == "Moderate-Severe" & osanew$location_name == location_name] * pop_female) / pop_both,
    PAFRR = ifelse(!is.na(RR), (osa_value * (RR - 1) / (osa_value * (RR - 1) + 1)), NA), 
    PAFOR = ifelse(!is.na(OR), paf_or(OR, prevalence, osa_value), NA),
    PAF = ifelse(is.na(PAFOR), PAFRR, PAFOR),
    ## Prevalents per conditions
    prevalent_cases = prevalence * pop_both, ## Taudin prevalenssi * populaatio, ok
    prevalent_cases_influenced_osa = PAF * prevalent_cases, ## PAF * prevalent_cases, ok
    ## Costs per conditions
    direct_cost = prevalent_cases_influenced_osa * annual_direct_healthcare_cost, ## ok
    direct_non_healthcare_cost = prevalent_cases_influenced_osa * annual_direct_nonhealthcare_cost, ## ok
    productivity_lost_cost = prevalent_cases_influenced_osa * annual_productivity_losses_cost, ## ok
    direct_cost = ifelse(is.na(direct_cost), 0 , direct_cost),
    direct_non_healthcare_cost = ifelse(is.na(direct_non_healthcare_cost), 0 , direct_non_healthcare_cost),
    productivity_lost_cost = ifelse(is.na(productivity_lost_cost), 0 , productivity_lost_cost),
    total_costs = direct_cost + direct_non_healthcare_cost + productivity_lost_cost) -> slapnea_cost1

## Cost per patient and sums
slapnea_cost1 %>% 
  group_by(location_name, pop_female, pop_male) %>% 
  summarise(
    ## Sums of costs
    direct_cost = sum(direct_cost, na.rm = T),
    direct_non_healthcare_cost = sum(direct_non_healthcare_cost, na.rm = T),
    productivity_lost_cost = sum(productivity_lost_cost, na.rm = T)
    )  %>% 
  mutate(
    ## Absolute values for dividing the cost per patient
    absolute_value_severe_moderate = ( (pop_female * osanew$rate[osanew$var == "Moderate" & osanew$gender == "Female" & osanew$location_name == location_name]) + (pop_female * osanew$rate[osanew$var == "Severe" & osanew$gender == "Female" & osanew$location_name == location_name]) + (pop_male * osanew$rate[osanew$var == "Moderate" & osanew$gender == "Male" & osanew$location_name == location_name]) + (pop_male * osanew$rate[osanew$var == "Severe" & osanew$gender == "Male" & osanew$location_name == location_name])), ## ok
    # absolute_value_mild = ( pop_female * osa$rate[osa$var == "Mild" & osa$gender == "Female"]) + (pop_male * osa$rate[osa$var == "Mild" & osa$gender == "Male"] ), ## ok
    ## Costs per patients
    patient_direct_cost = direct_cost / absolute_value_severe_moderate, ## ok
    patient_nonhealthcare_cost = direct_non_healthcare_cost / absolute_value_severe_moderate, ## ok
    patient_productivity_cost = productivity_lost_cost / absolute_value_severe_moderate, ## ok
    ## Total cost per patient
    patient_total_cost = patient_direct_cost + patient_nonhealthcare_cost + patient_productivity_cost ## ok
  ) -> slapnea22
## Write to calculation database
saveRDS(slapnea22, file = "data/slapnea22.RDS")

# View(slapnea_cost2[, c("location_name", "patient_total_cost")])
# Results: costs are almost right, 50 - 100 difference. WHYYYY???

slapnea22 %>% 
  left_join(money_correction, by = "location_name") %>% 
  mutate(
    index= ifelse(is.na(index), 1, index),
    direct_cost = direct_cost * index,
    direct_non_healthcare_cost = direct_non_healthcare_cost * index,
    productivity_lost_cost = productivity_lost_cost * index,
    total_cost = direct_cost + direct_non_healthcare_cost +productivity_lost_cost,
    ## Per patient
    patient_direct_cost = direct_cost / absolute_value_severe_moderate, ## ok
    patient_nonhealthcare_cost = direct_non_healthcare_cost / absolute_value_severe_moderate, ## ok
    patient_productivity_cost = productivity_lost_cost / absolute_value_severe_moderate, ## ok
    patient_total_cost = patient_direct_cost + patient_nonhealthcare_cost + patient_productivity_cost ## ok
  ) -> slapnea22_eurostat
## Write to calculation database
saveRDS(slapnea22_eurostat, file = "data/slapnea22_eurostat.RDS")

```

## Map data

```{r}
## Data for map
# Read this shape file with the rgdal library. 
library(rgdal)
world_spdf <- readOGR( 
  dsn= paste0(getwd(),"/data/map/") , 
  layer="TM_WORLD_BORDERS_SIMPL-0.3",
  verbose=FALSE
)

## Countries in our data
countries <- tbl(con, "summary_slapnea_costs") %>% 
  group_by(location_name) %>% 
  summarise(1) %>% 
  collect()
countries <- countries$location_name

## Countries in world data
sort(world_spdf$NAME)

## Rename few countries in world data and subset
world_spdf$NAME[world_spdf$NAME == "The former Yugoslav Republic of Macedonia"] <- "Macedonia"
world_spdf$NAME[world_spdf$NAME == "Republic of Moldova" ] <- "Moldova"
europe_spdf <- subset(world_spdf, world_spdf$NAME %in% countries)
countries[!countries %in% europe_spdf$NAME]

## Write to calculation database
saveRDS(europe_spdf, file = "data/europe_spdf.RDS")
```

## Save and copy data

```{r, eval=FALSE}
## Copy to sleep22visualizer
fils <- c(paste0(here::here(), "/data/europe_spdf.RDS"), paste0(here::here(), "/data/slapnea22.RDS"), paste0(here::here(), "/data/slapnea22_eurostat.RDS"))
fs::file_copy(fils, "/Users/japmiett/projects/sleep22visualizer/data/", overwrite = T)
```








## Building the calculator

## Results


## Thoughts


## References

Armeni et al. (2019) Cost-of-illness study of Obstructive Sleep Apnea Syndrome (OSAS) in Italy

<!--chapter:end:10-article.Rmd-->



## Calculate cost with varma cause_id
## All cause_id:t
## Check differences between costs

<!--chapter:end:test-cause_id.Rmd-->



## Unit check that PAF calculations are ok
## Use Italy data from article and calculate PAF

<!--chapter:end:test-paf_calc.Rmd-->

